Supplementary Information: Drought sensitivity is not a species trait: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-10-07 doc/response_to_reviewer_R1.Rmd

---
title: 'Responses to the Reviewer'
output: 
    bookdown::word_document2: 
    #   reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
      reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
suppress-bibliography: TRUE
---

Dear Editor,

We are pleased to submit a revised version of our manuscript, "Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments" () for consideration for publication in Ecology Letters. 
We have addressed all points raised by the reviewers as detailed in our responses to the reviewers. 
The most significant changes were discussing global implications as well as species effects in more detail. 
We have also revised the methodology and results to highlight the correspondence between the predictions and the results. 
We have also updated Figures 1 and 2 as well as the relevant figures in the supplementary material as requested by the reviewers.

Thank you for considering this revised version for publication. 
We look forward to your response.

Sincerely,
On behalf of all coauthors,
Krishna Anujan

\newpage

## Referees' comments to the author(s):

### Referee: 1

@@

**RESPONSE: Thank you for your favourable review of the manuscript and for the constructive comments. We have now revised the manuscript based on these suggestions and that of the other reviewer which have helped improve the mansucript. Our detailed comments are below.**

Main comments:

•	It is always helpful to know the study site. I recommend including some pictures of the forest and study area.

**RESPONSE: Thanks for this comment. This is a great idea, and we would like to include this as well to provide further context to readers. Given that we are at the limit of display items in the main text, we have included this as Figure S1 in the supplementary material.**

@@

•	276: This section is important, but it's not a relevant result; rather, it's a methodological aspect that allowed the study to be conducted. I think it should be included in the materials and methods section.

**RESPONSE: Thanks for this suggestion. We have now moved the description of the droughts to the methods section and have only retained a brief description of the differences among droughts in the results.**

•	Since the study is based on only one site, it is relevant to improve the discussion section with other already published study sites (add more references), mainly link the global implications, such mortality rate and vulnerability in tropical forests. See https://urldefense.proofpoint.com/v2/url?u=https-3A__doi.org_10.1126_science.adq6607&d=DwIFaQ&c=009klHSCxuh5AI1vNQzSO0KGjl4nbi2Q0M1QLJX9BeE&r=6fH468T58at0TtM59sMf109A_dRgo6XTfGMOVX5Ewrw&m=fPE_4y1UPsTl9St_Y44IUO7VIaQJOJIUiHtK15-Nr7g4UiVKbtYQ-TE0043SLAj6&s=SxqVZgFkbqAZ0-VEvdQntXWpGEUu9lr25rgxQtwaVdw&e=

**RESPONSE: Thanks for this suggestion. We were excited to see this global study published recently. We have now discussed our focused work in relation to this global study based on the alignment of the key message - the diversity of drought responses with drought types and species traits (deciduousness). However, dendroband datasets can be key to filling gaps in spatial and climatic coverage of tree rings records in the tropics (reviewed by @groenendijk_importance_2025). Moreover, our results showcasing response diversity among individuals within a species demonstrates an important caveat to aggregate/species-chronology-based inferences from tree-ring records.**

@@

•	There was something missing to discuss about the species in particular, some that have been more affected than others and why, especially those in conservation categories.

**RESPONSE: Thanks for this suggestion. We have now included a supplementary table (Table S1) that details the response of specific species to specific drought events. This summary highlights our main message that the species responses are dissimilar across drought events. This table also includes available information on the IUCN categories of the species analysed (accessed in September 2025) where information was avaialable. Two of the species analysed are "Endangered", including *Afzelia xylocarpa*, the species that was discussed. There are also "Vulnerable" and "Data Deficient" species in the list, some of which have consistent negative drought sensitivities. This table would therefore be useful for forest managers and conservation professionals and to guide species-specific research directions. We have also added a line highlighting the contrasts between species responses in the discussion, which now reads:**

**"An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig. S5, Table S1).
Extrapolating from these could lead to erroneous inferences on species with conservation or commercial significance, e.g. *Dipterocarpus alatus* which presents a different drought response (Table S1)."**

@@

**As to why different species might be responding differently, a key difference could be their evolutionary strategies along hydraulic safety vs. efficiency. 
We have discussed our results in light of deciduousness as an example.
However, we recognise that many other above and belowground traits may contribute to these differences and need to examined further.
We have edited text in the discussion to read:**

**"Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments."**

@@

**RESPONSE: Thanks for this comment. We have revised Figure 2 to remove species lines. Originally, we had chosen to include the top 10 species in the plot to illustrate both synchrony and divergence in species responses, but the results text reported on growth and sensitivity values across all species considered (n=30), shown in the Supplementary Appendix. We agree that this misalignment in species choice was confusing and have removed it from Figure 2 and referenced the relevant figure in the results text.**

![](display/Fig2_alt2.png)

339-346: I had a hard time understanding this result, on the one hand it says that Crown exposure was associated with greater growth declines in 2010, but not for 2015 and 2020. However, according Fig 5b all all crown exposure categories had negative sensitivities in 2015 and 2020.

**RESPONSE: We agree the wording could be clearer. Our focus is on how growth sensitivity changes with crown exposure. Since all crown exposure categories had negative growth sensitivities in 2015 and 2020, crown exposure was not associated with large differences in the magnitude or direction of sensitivity. In other words, more exposed trees had more or less similar growth declines compared to less exposed/understory trees. We have now clarified this in the text which now reads:**

**"Predicted effects of CII on sensitivity decreased monotonically, with strong decreases in 2010, weak decrease in 2015, and no effect in 2020."**

@@

Figure 1a: “Dy” is not correct

**RESPONSE: Thanks for noticing this error. We have fixed this to say "Day of year".**

Fig. 2. It is unnecessary to include the years 2023 and 2024 on the y-axis, as they were not evaluated. I think this figure could be analyzed further. For example, it took two years after 2014 for growth to recover to previous year (for ALL species) unlike in 2010 and 2020, when it rebounded within a year to levels equal to or higher than the previous year.

**RESPONSE: Thanks for this suggestion. We have removed 2023 and 2024 from the x-axis. While we agree that there is a lot more to discuss and analyse from this figure, we wanted to be careful not to overinterpret the growth patterns without adequate analysis of other covariates. For example, the pattern of suppressed growth after the 2015 drought is not necessarily because of lag times in recovery but more likely because drought conditions extended into 2016 - this was a multi-year drought - and El Nino conditions switched to La Nina only late in 2016, leading to a late monsoon. Our analysis was focussed on episodic droughts, and therefore we chose not to analyse and interpret the whole timeseries in this study. We recognise however, that lagged responses and compensatory responses would be important in such timeseries and hope to answer these as the dataset accumulates.**

@@

2 The topographic wetness index. Are there any measured soil moisture data in this plot? Since this index is only based on topography, it will be good if the author could provide some measured soil moisture data to calibrate the prediction based on model prediction.

**RESPONSE: Unfortunately soil moisture data from the plot is not available for the years of study across which dendroband data are available. HKK, being fairly remote, has limited access to power sources and other infrastructure that are needed to maintain such data collection. We are currently piloting a few basic soil moisture sensors, however, this will have limited value for calibration as they only measure the top layer of the soil and are limited in spatial scale compared to the spatial resolution of TWI data.**

3 In the case of deciduousness, the authors indicated that they used the deciduous index from Williams et al. (2008). Since different species might shed leaves in different season. Did the authors consider the impact of the timing of deciduousness on the growth?

**RESPONSE: Thanks for this interesting comment. We agree that there are several hydraulic strategies both above and belowground that might impact tree growth sensitivity to drought. We chose degree of deciduousness as a simple, illustrative example of a strategy that might be associated with some of this variation. We agree that the timing of deciduousness may be an important factor to consider, as might leaf turgor loss point, wood density etc. and considering these would be a key direction for future analyses, but beyond the scope of the current analysis.**

### 2025-10-09 doc/manuscript_hkk_drought_sensitivity.Rmd
updated manuscript file and track changes

**Number of Words in Abstract:** 146  
**Number of Words in Main Text:** 4999   
**Number of References:** 55  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

@@

# Statement of Authorship {-}
KA, SMM, SJD and KAT conceptualised the analysis. 
SB, SJD and NP have been leading long-term data collection at HKK, while HCM created and helped execute the dendrometer band protocol. 
KA cleaned the data and conducted formal analysis with support from SMM, SJD, HCM and KAT. 
KA wrote the original draft of the manuscript with support from KAT and SMM. 
All authors contributed to revision, review and editing of the manuscript.

@@

# Abstract {-}
Increased frequency and severity of droughts threatens forest health worldwide<!--, with important implications for this critical carbon sink-->. 
Tree species adaptations <!--to hot and dry conditions-->-- e.g., dry-season deciduousness in tropical seasonal forests -- and individual traits -- e.g., size, crown position -- shape drought resistance, but such resistance may be variable across species, microenvironments, and drought events. 
Here, we assess growth responses of 1820 trees across 30 species to three climatically distinct droughts in a seasonally dry tropical forest in Western Thailand. 
Species and individuals exhibited a wide range of growth responses within each drought, and differences in response intensity and affect among the drought events. 
Deciduous and evergreen species were more sensitive to wet- and dry-season drought respectively<!--, and wetter microclimates amplified this difference-->.
While individuals with more exposed crowns tended to grow less in all droughts, stem diameter and topographic wetness had variable effects.
<!-- Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines.  -->
Heterogeneous drought responses of species and individuals indicate potential biological insurance effects in diverse forests in the face of increased drought.

@@

Intense droughts are becoming more frequent [@ipcc_climate_2023], affecting the demography, diversity, and  carbon (C) cycling of forests globally. 
However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict [@evans_scaling_2025], particularly for tropical forests, <!--whose critical C sink is threatened by drought [@brando_droughts_2019] but--> for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, studies have separately shown that drought characteristics, species traits, tree size and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought. 
Specifically, if these factors act independently, tree drought responses may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought. 
In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove vulnerable to a drought with different characteristics. 
The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of climatic anomalies [@slette_how_2019].
<!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], differences in seasonal timing, duration, or intensity of droughts may result in variable tree growth responses, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry-season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
Tree species adaptations to drought vary widely, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
In tropical seasonal forests in particular, trees have evolved to optimize growth, reproduction, and survival (i.e., fitness) under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought vulnerability, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate $\pi_{tlp}$, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
<!-- Finally, the microenvironments that individual trees experience influence growth responses to drought leading to varied atmospheric and hydrological stresses with distinct ecological consequences. -->
Microenvironments defined by horizontal topographic and edaphic variation and vertical light and thermal variation further shape individual tree experience of, and response to, drought. 
While habitats with relatively moist soils (e.g., low-lying topography) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021].
Indeed, during drought, trees near streams experienced greater growth declines in a temperate forest [@mcgregor_tree_2021] and increased mortality in a tropical forest [@zuleta_droughtinduced_2017]. 
<!-- Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments. -->
Tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
Tree size can mediate microenvironments by influencing both crown exposure through tree height and belowground water access through rooting depth/size. 
For large trees, aboveground gradients in crown exposure could be partially offset by the tendency to have larger/deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and therefore the potential to access more reliable water sources [@stahl_depth_2013]. 
<!-- Sean::: I need to look at this source. Is this true? it doesn't really make sense unless coupled with other traits (e.g., deciduousness shifts accross ontogeny) as 'putting down a tap root' is one of the first things a seedling of a deep rooted species does (i.e., it's not allometry but strategy) -->
However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
While larger trees tend to suffer greater growth declines during drought [@bennett_larger_2015], this may be driven by crown exposure, by height <!--itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium --><!--, as described by Darcy's law -->and its pressures on stem water flow[@mcdowell_darcys_2015; @fernández-de-uña_role_2023], or by trait differences across vertical axis [@vinod_thermal_2023].
Thus, horizontal and vertical microenvironmental gradients and tree size modify tree drought responses, likely in interaction with drought characteristics and species traits.

@@

<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats. 
Specifically, we test predictions that:
i) both species and individuals respond differently to different droughts;
ii) on the species level, deciduous and evergreen species exhibit differential sensitivity depending on the timing of drought and topographic location;
iii) on the individual level, the influences of topographic location, crown exposure, and other size effects on drought responses are variable across droughts.
We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 
During this period, the forest experienced three different dry periods, two associated with ENSO events, and all three differing with respect to timing and duration (Table 1). 
With data on 1820 trees across 30 species, we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
We used a causal modelling framework to account for potential confounding relationships among the variables. 
This analysis adds important insight into the complexity of drought responses in species-rich seasonal dry tropical forests.

@@

In HKK, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following standard ForestGEO protocol [@muller-landau_metal_2008]. 
Selected were a size-stratified and spatially-stratified random sample, with additional trees chosen to fill out sample sizes for selected species of interest.
Thus relative abudances among the selected trees largely reflected abundance variation in the plot as a whole, with more individuals banded from the abundant species (n=1 to 230 trees per species). 
On each tree, the dendrometer band was installed at a height of 10 cm above the regular point of measurement for the whole-plot census meaning installation at 1.4 m on most trees, and at higher heights on buttressed trees with higher measurement points. 
Dendrobands were censused twice a year with the measurement periods typically centred in June and January. 
During each census, the size of the window opening along the band was measured using Mitutoyo digital callipers (precision = 0.01 mm) and the diameter of the tree at the dendrometer band was measured using standard DBH tape.

@@

We first  calculated the mean annual increment $\bar{a.inc_{i}}$ for each tree from annualised increment $ainc_{i,t}$.
We then calculated growth sensitivity for each tree $i$ for each drought year $y$ as 
$$
sensitivity_{i, t=y} = \frac{ainc_{i, t=y} - \bar{a.inc_{i}}}{\bar{a.inc_{i}}}
$$
By scaling the increment deviation to the mean annual growth of each tree, this metric allows us to compare across slow and fast growing individuals and species.

>>>>

**Number of Words in Abstract:** 146  
**Number of Words in Main Text:** 5054   
**Number of References:** 55  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

@@

# Statement of Authorship {-}
KA, SMM, SJD and KAT conceptualised the analysis. 
SB, SJD and NP have been leading long-term data collection at HKK, while HCM led the development of the dendrometer band protocol. 
KA cleaned the data and conducted formal analysis with support from SMM, SJD, HCM and KAT. 
KA wrote the original draft of the manuscript with support from KAT and SMM. 
All authors contributed to revision, review and editing of the manuscript.

@@

# Abstract {-}
Increased frequency and severity of droughts threatens forest health worldwide<!--, with important implications for this critical carbon sink-->. 
Tree species-specific adaptations <!--to hot and dry conditions-->-- e.g., dry-season deciduousness in tropical seasonal forests -- and individual traits -- e.g., size, crown position -- shape drought resistance, but such resistance may be variable across species, microenvironments, and drought events. 
Here, we assess growth responses of 1820 trees across 30 species to three climatically distinct droughts in a seasonally dry tropical forest in Western Thailand. 
Species and individuals exhibited a wide range of growth responses within each drought, and differences in response intensity and affect among the drought events. 
Deciduous and evergreen species were more sensitive to wet- and dry-season drought respectively<!--, and wetter microclimates amplified this difference-->.
While individuals with more exposed crowns tended to grow less in all droughts, stem diameter and topographic wetness had variable effects.
<!-- Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines.  -->
Heterogeneous drought responses of species and individuals indicate potential biological insurance effects in diverse forests in the face of increased drought.

@@

Intense droughts are becoming more frequent globally [@ipcc_climate_2023], affecting the demography, diversity, and  carbon (C) cycling of forests. 
However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict [@evans_scaling_2025], particularly for tropical forests, <!--whose critical C sink is threatened by drought [@brando_droughts_2019] but--> for which long-term records of annual woody tree growth are limited [@zuidema_tropical_2022;@groenendijk_importance_2025]. 
For forests worldwide, studies have separately shown that drought characteristics, species traits, tree size and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought. 
If these factors act independently, tree drought responses may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought. 
Alternatively, if these factors interact, then species and individuals that exhibit resistance to one drought may prove vulnerable to a drought with different characteristics. 
These complex and diverse drought responses may buffer ecosystem function even as intense droughts become more common and may act as a mechanism for maintaining species diversity [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
<!-- Sean::: This paragraph is a bit redundant and all over the place. 1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation. Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways. [Then go into some details with citations on ?] -->
Drought is notoriously hard to define and can be characterized based on the timing, severity, or duration of climatic anomalies in a variety of meteorological variables[@slette_how_2019].
<!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
<!-- Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], differences in seasonal timing, duration, or intensity of droughts may result in variable tree growth responses, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018]. -->
This difficulty in definition, and subsequent predictions of tree growth responses, is pronounced in seasonal forests with diverse adaptive strategies that have evolved under consistent low-rainfall periods [@albert_cryptic_2019;@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
In tropical forests with strong dry seasons, tree drought responses may depend less on the climatic extent of the drought and more on drought timing relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology]. It seems like these arguments are a bit redundant. Or meandering. ---->
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
For example, interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->has been shown to be most sensitive to dry-season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]. 
However, these analyses represent a limited set of tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

@@

<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
Species adaptations to drought vary widely, often leading to classification of tree species as "drought tolerant" or "drought sensitive" or "drought avoidant" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
<!-- In tropical seasonal forests in particular, trees optimize growth, reproduction, and survival (i.e., fitness) under dry seasons, along a continuum spanning intensive allocation to hydraulic safety or to hydraulic efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021]. -->
In tropical seasonal forests, many tree species vary in their hydraulic strategies along a continuum from drought tolerant species having high hydraulic safety, low hydraulic efficiency and high tissue investment to drought "avoidant" species with high hydraulic efficiency, low hydraulic safety and low investment to tissues [@gonzález-m._diverging_2021].
Traits associated with hydraulic safety and drought tolerance include deep roots (to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, to maintain gas exchange and photosynthesis under hot and dry conditions), and small vessels (to reduce embolism risk)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss [@desouza_drought_2020]. 
Under drought conditions, these strategies may confer more or less drought resistance depending on the specific trait combinations and drought characteristics [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
These strategies can also have developmental or ecophysiological controls.
For example, some species regulate $\pi_{tlp}$, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008;@hulshof_variation_2010].

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
<!-- Finally, the microenvironments that individual trees experience influence growth responses to drought leading to varied atmospheric and hydrological stresses with distinct ecological consequences. -->
Microenvironments defined by horizontal topographic and edaphic variation and vertical light and thermal variation shape individual tree experience of, and response to, drought. 
Although habitats with relatively moist soils (e.g., low-lying topography) are more hospitable during drought, trees in these microenvironments may be less drought-adapted [e.g., due to less negative $\pi_{tlp}$, @kunert_leaf_2021] and thus more vulnerable under severe droughts.
Indeed, during drought, trees near streams experienced greater growth declines in a temperate forest [@mcgregor_tree_2021] and increased mortality in a tropical forest [@zuleta_droughtinduced_2017]. 
<!-- Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments. -->
Tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019;@ma_tree_2023].
<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
Tree size can mediate microenvironments by influencing both crown exposure through tree height and belowground water access through rooting depth/size. 
For large trees, aboveground gradients in crown exposure could be partially offset by the tendency to have larger/deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and therefore the potential to access more reliable water sources [@stahl_depth_2013]. 
<!-- Sean::: I need to look at this source. Is this true? it doesn't really make sense unless coupled with other traits (e.g., deciduousness shifts accross ontogeny) as 'putting down a tap root' is one of the first things a seedling of a deep rooted species does (i.e., it's not allometry but strategy) -->
However, as with wet-adapted trees, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
Although larger trees tend to suffer greater growth declines during drought [@bennett_larger_2015], this may be driven by crown exposure, by height <!--itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium --><!--, as described by Darcy's law -->and its pressures on stem water flow[@mcdowell_darcys_2015;@fernández-de-uña_role_2023], or by systematic variation in leaf traits differences along the vertical axis [@vinod_thermal_2023].
Thus, horizontal and vertical microenvironmental gradients and tree size modify tree drought responses, likely in interaction with drought characteristics and species traits.

@@

<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats. 
We test predictions that:
i) both species and individuals respond differently to different droughts;
ii) on the species level, deciduous and evergreen species exhibit differential sensitivity depending on the timing of drought and topographic location;
iii) on the individual level, the influences of topographic location, crown exposure, and other size effects on drought responses are variable across droughts.
We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest in Thailand[@bunyavejchewin_seasonally_2011]. 
During this period, the forest experienced three different dry periods, two associated with ENSO events, and all three differing with respect to timing and duration (Table 1). 
With data on 1820 trees across 30 species, we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
We used a causal modelling framework to account for potential confounding relationships among the variables and to enable casual interpretation of estimated parameters. 
This analysis demonstrates the complexity of drought responses in species-rich seasonal dry tropical forests.

@@

In HKK, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following standard ForestGEO protocol [@muller-landau_metal_2008]. 
Most selected trees were part of a size-stratified and spatially-stratified random sample, with additional trees chosen to fill out sample sizes for species of interest.
Thus, relative abudances among the selected trees largely reflected abundance variation in the plot as a whole, with more individuals banded from the abundant species (n=1 to 230 trees per species). 
On each tree, the dendrometer band was installed at a height of 10 cm above the regular point of measurement for the whole-plot census meaning installation at 1.4 m on most trees, and at higher heights on buttressed trees with higher measurement points. 
Dendrobands were censused twice a year with the measurement periods typically centred in June and January. 
During each census, the size of the window opening along the band was measured using Mitutoyo digital callipers (precision = 0.01 mm) and the diameter of the tree at the dendrometer band was measured using standard DBH tape.

@@

We first  calculated the mean annual increment $\overline{a.inc_{i}}$ for each tree from annualised increment $ainc_{i,t}$.
We then calculated growth sensitivity for each tree $i$ for each drought year $y$ as 
$$
sensitivity_{i, t=y} = \frac{ainc_{i, t=y} - \overline{a.inc_{i}}}{\overline{a.inc_{i}}}
$$
By scaling the increment deviation to the mean annual growth of each tree, this metric allows us to compare across slow and fast-growing individuals and species.
